---
title: "Blood Data Analyses"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

# Overview of aims

What is the effect of Prebiotin vs placebo on blood measures (controlling for diet, age, ethnicity, stress?)
Did the intervention effect? 
  


```{r data, include=FALSE, echo=FALSE}
# load packages
source("code/load_packages.R")

# get data
source("code/get_cleaned_data.R")

theme_set(theme_bw())
pal = "Set1"
scale_colour_discrete <-  function(palname=pal, ...){
  scale_colour_brewer(palette=palname, ...)
}
scale_fill_discrete <-  function(palname=pal, ...){
  scale_fill_brewer(palette=palname, ...)
}

knitr::opts_chunk$set(out.width = "200%")

```

# Data

```{r}

blood_data <- read_excel("data/Blood measures data/Copy of Fiber Study Blood Results.xlsx")

meta_data <- microbiome_data$meta.dat

keepVar <- c("SubjectID", "Week", "Intervention", "Stress.Scale", "Ethnicity", "Gender", "Age")
meta_data <- meta_data [, keepVar] %>%
  filter(Week == 1)

mydata <- full_join(blood_data, meta_data)

# recoding
mydata <- mydata %>%
  mutate(female = ifelse(Gender == "F", 1, 0),
         c.age = Age - mean(Age),
         IntB = ifelse(Intervention == "B", 1, 0),
         Post = ifelse(time == "Post", 1, 0),
         c.stress = Stress.Scale - mean(Stress.Scale),
         hispanic = ifelse(Ethnicity %in% c("White", "Asian", "Native America"), 1, 0))

# for plotting
plot.data <- mydata[, c(colnames(mydata)[c(1:18,20:24)])]

plot.data <- plot.data %>%
  pivot_longer(cols=colnames(mydata)[3:18],
               names_to = "Variable",
               values_to = "value")

```

## Data summary

### Summary Statistics

```{r data-sum}

varNames <- c(colnames(mydata)[c(2:18,20)])

sum.dat <- mydata[,varNames[2:17]] %>%
  summarise_all(list(Mean=mean, SD=sd,
                     min=min, Median=median, Max=max))
sum.dat <- data.frame(matrix(unlist(sum.dat), ncol=5))
colnames(sum.dat) <- c("Mean", "SD", "Min", "Median", "Max")
rownames(sum.dat) <- varNames[2:17]

kable(sum.dat, format="html", digits=3)%>%
  kable_styling(full_width = T)

```

### Summary of these data by Intervention Group

```{r data-sum1}

varNames <- colnames(mydata)[c(20,3:18)]
sum.dat <- mydata[,varNames] %>%
  group_by(Intervention) %>%
  summarise_all(list(Mean=mean, SD=sd,
                     min=min, Median=median, Max=max))

a <- data.frame(matrix(unlist(sum.dat[1,-1]), ncol=5))
b <- data.frame(matrix(unlist(sum.dat[2,-1]), ncol=5))
a <- cbind(rep("A", 16), a); colnames(a) <- c("Intervention", "Mean", "SD", "Min", "Median", "Max")
b <- cbind(rep("B", 16), b); colnames(b) <- c("Intervention", "Mean", "SD", "Min", "Median", "Max")

sum.dat <- rbind(a,b)
sum.dat <- data.frame(Variable=rep(varNames[-1],2), sum.dat)
 
sum.dat <- arrange(sum.dat, Variable)
kable(sum.dat, format="html", digits=3)%>%
  kable_styling(full_width = T)

```

### Summary of these data by Pre-Post

```{r data-sum2}

varNames <- colnames(mydata)[c(2:18)]

sum.dat <- mydata[,varNames] %>%
  group_by(time) %>%
  summarise_all(list(Mean=mean, SD=sd,
                     min=min, Median=median, Max=max))

a <- data.frame(matrix(unlist(sum.dat[1,-1]), ncol=5))
b <- data.frame(matrix(unlist(sum.dat[2,-1]), ncol=5))
a <- cbind(rep("Post", 16), a); colnames(a) <- c("Pre-Post", "Mean", "SD", "Min", "Median", "Max")
b <- cbind(rep("Pre", 16), b); colnames(b) <- c("Pre-Post", "Mean", "SD", "Min", "Median", "Max")

sum.dat <- rbind(b,a)
sum.dat <- data.frame(Variable=rep(varNames[-1],2), sum.dat)
 
sum.dat <- arrange(sum.dat, Variable)
kable(sum.dat, format="html", digits=3)%>%
  kable_styling(full_width = T)

```

### Summary of these data by Intervention & Pre-Post

```{r data-sum3}

varNames <- colnames(mydata)[c(20, 2:18)]

sum.dat <- mydata[,varNames] %>%
  group_by(Intervention, time) %>%
  summarise_all(list(Mean=mean, SD=sd,
                     min=min, Median=median, Max=max))

a2 <- data.frame(matrix(unlist(sum.dat[1,-c(1:2)]), ncol=5))
a1 <- data.frame(matrix(unlist(sum.dat[2,-c(1:2)]), ncol=5))
b1 <- data.frame(matrix(unlist(sum.dat[3,-c(1:2)]), ncol=5))
b2 <- data.frame(matrix(unlist(sum.dat[4,-c(1:2)]), ncol=5))

a2 <- cbind(rep("A", 16), rep("Post", 16), a2); colnames(a2) <- c("Intervention", "Pre-Post", "Mean", "SD", "Min", "Median", "Max")
a1 <- cbind(rep("A", 16), rep("Pre", 16), a1); colnames(a1) <- c("Intervention", "Pre-Post", "Mean", "SD", "Min", "Median", "Max")

b1 <- cbind(rep("B", 16), rep("Pre", 16), b1); colnames(b1) <- c("Intervention", "Pre-Post", "Mean", "SD", "Min", "Median", "Max")

b2 <- cbind(rep("B", 16), rep("Post", 16), b2); colnames(b2) <- c("Intervention", "Pre-Post", "Mean", "SD", "Min", "Median", "Max")

sum.dat <- rbind(a1,a2, b1, b2)
sum.dat <- data.frame(Variable=rep(varNames[-c(1:2)],4), sum.dat)
 
sum.dat <- arrange(sum.dat, Variable)
kable(sum.dat, format="html", digits=3)%>%
  kable_styling(full_width = T)

```


# Analysis

Next, the aim is to more formally test for differences between intervention groups.


```{r fig.height=12}

# plot
p1 <- ggplot(plot.data,
             aes(x=time, y=value, group=time, color=time))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width = 0.25)+
  facet_wrap(.~Variable, scales = "free") +
  labs(x=NULL, title="Pre and Post Blood Measures")+
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=10))
p1

```

```{r fig.height=20}

p2 <- ggplot(plot.data,
             aes(x=time, y=value, group=time, color=time))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width = 0.25, size=2)+
  labs(x=NULL,y=NULL, title="Blood Measures by Time & Intervention Group")+
  facet_grid(Variable~Intervention, scale="free", labeller = labeller(Intervention = c(A = "Group A", B = "Group B")))+
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=10),
        strip.text = element_text(size=10))
p2


```

## Run Analyses

This is simply to double check that no differences occured at baseline.


```{r}

results.out <- list()

i <- 1
varNames <- colnames(mydata)[3:18]

for(i in 1:length(varNames)){
  cat("\n\n=======================")
  cat("\n=======================")
  cat("\nOutcome:\t",varNames[i])
  cat("\n")
  form <- as.formula(paste0(varNames[i], "~ IntB + Post + IntB:Post + c.age"))
  fit <- lm(form, mydata)
  anova(fit)
  
  results.out[[varNames[i]]] <- summary(fit)[["coefficients"]]
  
  
  # diagnostic plots
  layout(matrix(c(1,2,3,4),2,2)) # optional layout
  plot(fit) # diagnostic plots
  
  # normality again
  shapiro.test(residuals(fit))
  
  # independence
  durbinWatsonTest(fit)
  layout(matrix(c(1),1,1))
  acf(residuals(fit))
  
  # nice wrapper function to generally test a lot of stuff
  gvmodel <- gvlma(fit)
  summary(gvmodel)

}


```

## Results Adjusted for Multiple Comparisons

```{r echo=F}

unlist.res <- function(dat, ...){
  #dat<- results.out
  nc <- ncol(dat[[1]])
  cnm <- colnames(dat[[1]])
  rnm <- names(dat)
  
  # unlist
  res <- data.frame(matrix(unlist(dat[[1]]), ncol=nc))
  colnames(res) <- colnames(dat[[1]])
  res$Outcome <- rep(rnm[1], nrow(res))
  res$Parameter <- rownames(dat[[1]])
  for(i in 2:length(rnm)){
    resi <- data.frame(matrix(unlist(dat[[i]]), ncol=nc))
    colnames(resi) <- colnames(dat[[i]])
    resi$Outcome <- rep(rnm[i], nrow(resi))
    resi$Parameter <- rownames(dat[[i]])
    res <- full_join(res, resi)
  }
  
  return(res)
}


```


```{r}


all_results <- unlist.res(results.out)

all_results <- filter(all_results, Parameter %in% c("IntB", "Post", "IntB:Post"))
all_results$p.adj <- p.adjust(all_results$`Pr(>|t|)`, "fdr") 
all_results$sigflag <- ifelse(all_results$p.adj < 0.05, "*", " ")
all_results$`Pr(>|t|)` <- round(all_results$`Pr(>|t|)`, 3)
all_results$p.adj.fdr <- round(all_results$p.adj, 3)
all_results <- all_results[, c(5:6, 1:4, 7:8)]

kable(all_results, format="html", digits=3) %>%
  kable_styling(full_width = T)


#readr::write_csv(all_results, "tab/results_blood_data.csv")


```



