---
title: "Beta Diversity Analysis"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

# Beta-Diversity



```{r data, include=FALSE, echo=FALSE}
# load packages
source("code/load_packages.R")

# get data
source("code/get_cleaned_data.R")

theme_set(theme_bw())
pal = "Set1"
scale_colour_discrete <-  function(palname=pal, ...){
  scale_colour_brewer(palette=palname, ...)
}
scale_fill_discrete <-  function(palname=pal, ...){
  scale_fill_brewer(palette=palname, ...)
}

knitr::opts_chunk$set(out.width = "200%")

```


# General Ordination Plot

```{r}

# Ordinate
pcoa <- ordinate(
  physeq = phylo_data, 
  method = "PCoA", 
  distance = "bray"
)

# Plot 
p <- plot_ordination(
  physeq = phylo_data,
  ordination = pcoa,
  color = "Week",
  shape = "Intervention",
  title = "PCoA plot of community differences over time"
) 
p +
  geom_point(aes(color = Week), alpha = 0.7, size = 4) +
  scale_color_manual(values = c("#a65628", "red", "#ffae19","#4daf4a")) +
  geom_point(colour = "grey90", size = 1.5) 

```




# PERMANOVA

Below, we present the permutation based ANOVA results for community differences.
We conducted these analyses with three models

1. Using Week
2. Using Intervention
3. Using Week $\times$ Intervention interaction

Compute the distances matrices and get ordination objects.

```{r}

# Calculate distance matrices
dist_bray <- phyloseq::distance(phylo_data, method = "bray")
dist_unwt <- phyloseq::distance(phylo_data, method="unifrac", weighted=F)
dist_wt <- phyloseq::distance(phylo_data, method="unifrac", weighted=T)

# plot ordination
ord_bray = ordinate(phylo_data, method="PCoA", distance=dist_bray)
ord_unwt = ordinate(phylo_data, method="PCoA", distance=dist_unwt)
ord_wt = ordinate(phylo_data, method="PCoA", distance=dist_wt)

# sample data
df <- data.frame(sample_data(phylo_data))

# colors
cols <- c("#fe9700","#00a2f2", "#662a00", "#c91acb","grey60","#858c69", "#a8863a", "#737373", "#d43f1f", "#5dd047",  "#ffff59")

```

## Analysis by Week

```{r fig1, fig.height = 8, fig.width = 4, fig.align = "center"}

p1 <- plot_ordination(phylo_data, ord_bray, color="Week") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  labs(title="PCoA on Bray-Curtis distance, by Week")

p2 <- plot_ordination(phylo_data, ord_unwt, color="Week") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  labs(title="PCoA on Unweighted Unifrac distance, by Week")

p3 <- plot_ordination(phylo_data, ord_wt, color="Week") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  labs(title="PCoA on Weight Unifrac distance, by Week")
p1 + p2 + p3 + plot_layout(ncol=1)

# Bray-Curtis
adonis(dist_bray ~ Week, data = df)
beta <- betadisper(dist_bray, df$Week)
permutest(beta)

# Unweighted Unifrac
adonis(dist_unwt ~ Week, data = df)
beta <- betadisper(dist_unwt, df$Week)
permutest(beta)

# Weighted Unifrac
adonis(dist_wt ~ Week, data = df)
beta <- betadisper(dist_wt, df$Week)
permutest(beta)

```

## Analysis by Intervention

```{r fig2, fig.height = 8, fig.width = 4, fig.align = "center"}


p1 <- plot_ordination(phylo_data, ord_bray, color="Intervention") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  labs(title="PCoA on Bray-Curtis distance, by Intervention")

p2 <- plot_ordination(phylo_data, ord_unwt, color="Intervention") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  labs(title="PCoA on Unweighted Unifrac distance, by Intervention")

p3 <- plot_ordination(phylo_data, ord_wt, color="Intervention") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  labs(title="PCoA on Weight Unifrac distance, by Intervention")
p1 + p2 + p3 + plot_layout(ncol=1)

# Bray-Curtis
adonis(dist_bray ~ Intervention, data = df)
beta <- betadisper(dist_bray, df$Intervention)
permutest(beta)

# Unweighted Unifrac
adonis(dist_unwt ~ Intervention, data = df)
beta <- betadisper(dist_unwt, df$Intervention)
permutest(beta)

# Weighted Unifrac
adonis(dist_wt ~ Intervention, data = df)
beta <- betadisper(dist_wt, df$Intervention)
permutest(beta)

```


## Analysis by Intervention

```{r fig3, fig.height = 8, fig.width = 4, fig.align = "center"}

p1 <- plot_ordination(phylo_data, ord_bray,
                      color="Week", shape="Intervention") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  geom_point(colour = "grey90", size = 1.5) +
  labs(title="PCoA on Bray-Curtis distance, by Week & Intervention")+
  theme(legend.position = "none")

p2 <- plot_ordination(phylo_data, ord_unwt,
                      color="Week", shape="Intervention") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  geom_point(colour = "grey90", size = 1.5) +
  labs(title="PCoA on Unweighted Unifrac distance, by Week & Intervention")+
  theme(legend.position = "none")

p3 <- plot_ordination(phylo_data, ord_wt,
                      color="Week", shape="Intervention") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  geom_point(colour = "grey90", size = 1.5) +
  labs(title="PCoA on Weight Unifrac distance, by Week & Intervention")+
  theme(legend.position = "bottom")
p1 + p2 + p3 + plot_layout(ncol=1)

# Bray-Curtis
adonis(dist_bray ~ Intervention*Week, data = df)
beta <- betadisper(dist_bray, interaction(df$Intervention, df$Week))
permutest(beta)

# Unweighted Unifrac
adonis(dist_unwt ~ Intervention*Week, data = df)
beta <- betadisper(dist_unwt, interaction(df$Intervention, df$Week))
permutest(beta)

# Weighted Unifrac
adonis(dist_wt ~ Intervention*Week, data = df)
beta <- betadisper(dist_wt, interaction(df$Intervention, df$Week))
permutest(beta)

```

