---
title: "Modeling Changes in Alpha Diversity over Time"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

# Overview of aims

What is the effect of Prebiotin vs placebo on anthropometrics (controlling for diet, age, ethnicity, stress?)
Did the intervention mitigate excess weight gain? 
  
  * BMI 
  * Lean Body Mass 
  * Visceral Fat Level 
  * Weight

Therefore, data are analyzed at the last timepoint observed for each participant (week 12).
If a participant is missing an outcome at week 12, the last observed score will be carried forward.

```{r data, include=FALSE, echo=FALSE}
# load packages
source("code/load_packages.R")

# get data
source("code/get_cleaned_data.R")

theme_set(theme_bw())
pal = "Set1"
scale_colour_discrete <-  function(palname=pal, ...){
  scale_colour_brewer(palette=palname, ...)
}
scale_fill_discrete <-  function(palname=pal, ...){
  scale_fill_brewer(palette=palname, ...)
}

knitr::opts_chunk$set(out.width = "200%")

```


# Analyses

What is the effect of Prebiotin vs. placebo on ...

```{r data-manipulate}

mydata <- microbiome_data$meta.dat
mydata <- arrange(mydata, desc(Week))
mydata <- distinct(mydata, SubjectID, .keep_all = T)
varNames <- c("SubjectID", "Week", "Age", "Ethnicity",
              "Gender", "Intervention", 
              "Height_cm", "Weight_pre", "Weight_post", 
              "LBM_pre", "LBM_post", "Visceral_Fat_Level_pre",
              "Visceral_Fat_Level_post", "Stress.Scale")
mydata <- mydata[, varNames]

mydata<- mydata %>%
  mutate(Weight_diff = Weight_pre - Weight_post,
         BMI_pre = Weight_pre/((Height_cm/100)**2),
         BMI_post = Weight_post/((Height_cm/100)**2),
         BMI_diff = BMI_pre - BMI_post,
         VFL_pre = Visceral_Fat_Level_pre,
         VFL_post = Visceral_Fat_Level_post,
         VFL_diff = VFL_pre - VFL_post,
         LBM_diff = LBM_pre - LBM_post,
         intB = ifelse(Intervention=="B", 1,0),
         c.age = Age - mean(Age),
         c.stress = Stress.Scale - mean(Stress.Scale),
         female = ifelse(Gender == "F", 1, 0),
         hispanic = ifelse(Ethnicity %in% c("White", "Asian", "Native America"), 1, 0))

plot.data <- mydata[,c(varNames[c(1:6,14,7:11)],  "Weight_diff", "BMI_pre", "BMI_post", "BMI_diff", "VFL_pre", "VFL_post", "VFL_diff", "LBM_diff")]
plot.data <- plot.data %>%
  pivot_longer(cols=c("Weight_pre", "Weight_post", "Weight_diff", "LBM_pre","LBM_post", "LBM_diff","BMI_pre", "BMI_post", "BMI_diff", "VFL_pre",  "VFL_post", "VFL_diff"),
               names_to = "Variable",
               values_to = "value")


```

Summary of these data is below

```{r data-sum}

varNames <- c("intB", "female", "hispanic", "Age", "Stress.Scale",
              "Weight_pre", "Weight_post", "Weight_diff",
              "BMI_pre", "BMI_post", "BMI_diff",
              "LBM_pre", "LBM_post", "LBM_diff",
              "VFL_pre", "VFL_post", "VFL_diff")
sum.dat <- mydata[,varNames] %>%
  summarise_all(list(Mean=mean, SD=sd,
                     min=min, Median=median, Max=max))
sum.dat <- data.frame(matrix(unlist(sum.dat), ncol=5))
colnames(sum.dat) <- c("Mean", "SD", "Min", "Median", "Max")
rownames(sum.dat) <- varNames

kable(sum.dat, format="html", digits=3)%>%
  kable_styling(full_width = T)

```

Summary of these data by Intervention Group

```{r data-sum2}

varNames <- c("Intervention",
              "female", "hispanic", "Age", "Stress.Scale",
              "Weight_pre", "Weight_post", "Weight_diff",
              "BMI_pre", "BMI_post", "BMI_diff",
              "LBM_pre", "LBM_post", "LBM_diff",
              "VFL_pre", "VFL_post", "VFL_diff")
sum.dat <- mydata[,varNames] %>%
  group_by(Intervention) %>%
  summarise_all(list(Mean=mean, SD=sd,
                     min=min, Median=median, Max=max))

a <- data.frame(matrix(unlist(sum.dat[1,-1]), ncol=5))
b <- data.frame(matrix(unlist(sum.dat[2,-1]), ncol=5))
a <- cbind(rep("A", 16), a); colnames(a) <- c("Intervention", "Mean", "SD", "Min", "Median", "Max")
b <- cbind(rep("B", 16), b); colnames(b) <- c("Intervention", "Mean", "SD", "Min", "Median", "Max")

sum.dat <- rbind(a,b)
sum.dat <- data.frame(Variable=rep(varNames[-1],2), sum.dat)
 
sum.dat <- arrange(sum.dat, Variable)
kable(sum.dat, format="html", digits=3)%>%
  kable_styling(full_width = T)

```

Next, the aim is to more formally test for differences between intervention groups.

```{r}

results.diff <- list()

```

## BMI 

```{r}

# plot
p1 <- ggplot(filter(plot.data, Variable %like% "BMI"),
             aes(x=Variable, y=value))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width = 0.25)+
  scale_x_discrete(labels=c("BMI_pre"="Pre", "BMI_post"="Post"),
                   limits=c("BMI_pre", "BMI_post"))+
  labs(x=NULL, y="BMI", title="Pre and Post BMI")+
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=10))
#p1

p2 <- ggplot(filter(plot.data, Variable %like% "BMI"),
             aes(x=Variable, y=value))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width = 0.25, size=2)+
  scale_x_discrete(labels=c("BMI_pre"="Pre", "BMI_post"="Post"),
                   limits=c("BMI_pre", "BMI_post"))+
  labs(x=NULL, y="BMI", title="Pre and Post BMI by Intervention Group")+
  facet_grid(.~Intervention, labeller = labeller(Intervention = c(A = "Group A", B = "Group B")))+
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=10),
        strip.text = element_text(size=10))
#p2
p <- p1 + p2
p

```

### Pre BMI

This is simply to double check that no differences occured at baseline.


```{r}

# Pre INtervention BMI - should beno difference
fit <- lm(BMI_pre ~ Intervention + c.age + hispanic + c.stress, mydata)
anova(fit)

# diagnostic plots
layout(matrix(c(1,2,3,4),2,2)) # optional layout
plot(fit) # diagnostic plots

# normality again
shapiro.test(residuals(fit))

# independence
durbinWatsonTest(fit)
layout(matrix(c(1),1,1))
acf(residuals(fit))

# nice wrapper function to generally test a lot of stuff
gvmodel <- gvlma(fit)
summary(gvmodel)


```

No significant differences.

### Post BMI

Test for significant differences after intervention

```{r}

# Post INtervention BMI - should beno difference
# not controls
fit <- lm(BMI_post ~ Intervention, mydata)
anova(fit)
summary(fit)

fit <- lm(BMI_post ~ Intervention + c.age + hispanic + c.stress,
          mydata)
anova(fit)
summary(fit)

# diagnostic plots
layout(matrix(c(1,2,3,4),2,2)) # optional layout
plot(fit) # diagnostic plots

# normality again
shapiro.test(residuals(fit))

# independence
durbinWatsonTest(fit)
layout(matrix(c(1),1,1))
acf(residuals(fit))

# nice wrapper function to generally test a lot of stuff
gvmodel <- gvlma(fit)
summary(gvmodel)


```

No significant difference found.

### Difference Scores of BMI

Test for significant differences after intervention

```{r}

# Post INtervention BMI - should beno difference
# not controls
fit <- lm(BMI_diff ~ Intervention, mydata)
anova(fit)
summary(fit)

fit <- lm(BMI_diff ~ Intervention + c.age + hispanic + c.stress,
          mydata)
anova(fit)
summary(fit)
results.diff[["BMI"]] <- summary(fit)[["coefficients"]]

# diagnostic plots
layout(matrix(c(1,2,3,4),2,2)) # optional layout
plot(fit) # diagnostic plots

# normality again
shapiro.test(residuals(fit))

# independence
durbinWatsonTest(fit)
layout(matrix(c(1),1,1))
acf(residuals(fit))

# nice wrapper function to generally test a lot of stuff
gvmodel <- gvlma(fit)
summary(gvmodel)


```

Therefore, we have evidence that BMI is not effected by Prebiotic vs. placebo.

## Lean Body Mass (LBM)

```{r}

# plot
p1 <- ggplot(filter(plot.data, Variable %like% "LBM"),
             aes(x=Variable, y=value))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width = 0.25, size=2)+
  scale_x_discrete(labels=c("LBM_pre"="Pre", "LBM_post"="Post"),
                   limits=c("LBM_pre", "LBM_post"))+
  labs(x=NULL, y="LBM", title="Pre and Post LBM")+
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=10))
#p1

p2 <- ggplot(filter(plot.data, Variable %like% "LBM"),
             aes(x=Variable, y=value))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width = 0.25, size=2)+
  scale_x_discrete(labels=c("LBM_pre"="Pre", "LBM_post"="Post"),
                   limits=c("LBM_pre", "LBM_post"))+
  labs(x=NULL, y="LBM", title="Pre and Post LBM by Intervention Group")+
  facet_grid(.~Intervention, labeller = labeller(Intervention = c(A = "Group A", B = "Group B")))+
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=10),
        strip.text = element_text(size=10))
#p2
p <- p1 + p2
p

```

### Pre LBM

This is simply to double check that no differences occured at baseline.


```{r}

# Pre INtervention LBM - should beno difference
fit <- lm(LBM_pre ~ Intervention + c.age + hispanic + c.stress, mydata)
anova(fit)

# diagnostic plots
layout(matrix(c(1,2,3,4),2,2)) # optional layout
plot(fit) # diagnostic plots

# normality again
shapiro.test(residuals(fit))

# independence
durbinWatsonTest(fit)
layout(matrix(c(1),1,1))
acf(residuals(fit))

# nice wrapper function to generally test a lot of stuff
gvmodel <- gvlma(fit)
summary(gvmodel)


```

No significant differences.

### Post LBM

Test for significant differences after intervention

```{r}

# Post INtervention LBM - should beno difference
# not controls
fit <- lm(LBM_post ~ Intervention, mydata)
anova(fit)
summary(fit)

fit <- lm(LBM_post ~ Intervention + c.age + hispanic + c.stress,
          mydata)
anova(fit)
summary(fit)

# diagnostic plots
layout(matrix(c(1,2,3,4),2,2)) # optional layout
plot(fit) # diagnostic plots

# normality again
shapiro.test(residuals(fit))

# independence
durbinWatsonTest(fit)
layout(matrix(c(1),1,1))
acf(residuals(fit))

# nice wrapper function to generally test a lot of stuff
gvmodel <- gvlma(fit)
summary(gvmodel)


```

### Difference Scores of LBM

Test for significant differences after intervention

```{r}

# Post INtervention LBM - should beno difference
# not controls
fit <- lm(LBM_diff ~ Intervention, mydata)
anova(fit)
summary(fit)

fit <- lm(LBM_diff ~ Intervention + c.age + hispanic + c.stress,
          mydata)
anova(fit)
summary(fit)
results.diff[["LBM"]] <- summary(fit)[["coefficients"]]

# diagnostic plots
layout(matrix(c(1,2,3,4),2,2)) # optional layout
plot(fit) # diagnostic plots

# normality again
shapiro.test(residuals(fit))

# independence
durbinWatsonTest(fit)
layout(matrix(c(1),1,1))
acf(residuals(fit))

# nice wrapper function to generally test a lot of stuff
gvmodel <- gvlma(fit)
summary(gvmodel)


```

Therefore, we have evidence that LBM is not effected by Prebiotic vs. placebo.


## Visceral Fat Level (VFL) Fat Mass


```{r}

# plot
p1 <- ggplot(filter(plot.data, Variable %like% "VFL"),
             aes(x=Variable, y=value))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width = 0.25, size=2)+
  scale_x_discrete(labels=c("VFL_pre"="Pre", "VFL_post"="Post"),
                   limits=c("VFL_pre", "VFL_post"))+
  labs(x=NULL, y="VFL", title="Pre and Post VFL")+
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=10))
#p1

p2 <- ggplot(filter(plot.data, Variable %like% "VFL"),
             aes(x=Variable, y=value))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width = 0.25, size=2)+
  scale_x_discrete(labels=c("VFL_pre"="Pre", "VFL_post"="Post"),
                   limits=c("VFL_pre", "VFL_post"))+
  labs(x=NULL, y="VFL", title="Pre and Post VFL by Intervention Group")+
  facet_grid(.~Intervention, labeller = labeller(Intervention = c(A = "Group A", B = "Group B")))+
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=10),
        strip.text = element_text(size=10))
#p2
p <- p1 + p2
p

```

### Pre VFL

This is simply to double check that no differences occured at baseline.


```{r}

# Pre INtervention VFL - should beno difference
fit <- lm(VFL_pre ~ Intervention + c.age + hispanic + c.stress, mydata)
anova(fit)

# diagnostic plots
layout(matrix(c(1,2,3,4),2,2)) # optional layout
plot(fit) # diagnostic plots

# normality again
shapiro.test(residuals(fit))

# independence
durbinWatsonTest(fit)
layout(matrix(c(1),1,1))
acf(residuals(fit))

# nice wrapper function to generally test a lot of stuff
gvmodel <- gvlma(fit)
summary(gvmodel)


```

No significant differences.

### Post VFL

Test for significant differences after intervention

```{r}

# Post INtervention VFL - should beno difference
# not controls
fit <- lm(VFL_post ~ Intervention, mydata)
anova(fit)
summary(fit)

fit <- lm(VFL_post ~ Intervention + c.age + hispanic + c.stress,
          mydata)
anova(fit)
summary(fit)

# diagnostic plots
layout(matrix(c(1,2,3,4),2,2)) # optional layout
plot(fit) # diagnostic plots

# normality again
shapiro.test(residuals(fit))

# independence
durbinWatsonTest(fit)
layout(matrix(c(1),1,1))
acf(residuals(fit))

# nice wrapper function to generally test a lot of stuff
gvmodel <- gvlma(fit)
summary(gvmodel)


```

### Difference Scores of VFL

Test for significant differences after intervention

```{r}

# Post INtervention VFL - should beno difference
# not controls
fit <- lm(VFL_diff ~ Intervention, mydata)
anova(fit)
summary(fit)

fit <- lm(VFL_diff ~ Intervention + c.age + hispanic + c.stress,
          mydata)
anova(fit)
summary(fit)
results.diff[["VFL"]] <- summary(fit)[["coefficients"]]

# diagnostic plots
layout(matrix(c(1,2,3,4),2,2)) # optional layout
plot(fit) # diagnostic plots

# normality again
shapiro.test(residuals(fit))

# independence
durbinWatsonTest(fit)
layout(matrix(c(1),1,1))
acf(residuals(fit))

# nice wrapper function to generally test a lot of stuff
gvmodel <- gvlma(fit)
summary(gvmodel)


```

Therefore, we have evidence that LBM is not effected by Prebiotic vs. placebo.

## Weight (kg)

```{r}

# plot
p1 <- ggplot(filter(plot.data, Variable %like% "Weight"),
             aes(x=Variable, y=value))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width = 0.25, size=2)+
  scale_x_discrete(labels=c("Weight_pre"="Pre", "Weight_post"="Post"),
                   limits=c("Weight_pre", "Weight_post"))+
  labs(x=NULL, y="Weight (kg)", title="Pre and Post Weight (kg)")+
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=10))
#p1

p2 <- ggplot(filter(plot.data, Variable %like% "Weight"),
             aes(x=Variable, y=value))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width = 0.25, size=2)+
  scale_x_discrete(labels=c("Weight_pre"="Pre", "Weight_post"="Post"),
                   limits=c("Weight_pre", "Weight_post"))+
  labs(x=NULL, y="Weight (kg)", title="Pre and Post Weight (kg) by Intervention Group")+
  facet_grid(.~Intervention, labeller = labeller(Intervention = c(A = "Group A", B = "Group B")))+
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=10),
        strip.text = element_text(size=10))
#p2
p <- p1 + p2
p

```

### Pre Weight

This is simply to double check that no differences occured at baseline.


```{r}

# Pre INtervention Weight - should beno difference
fit <- lm(Weight_pre ~ Intervention + c.age + hispanic + c.stress, mydata)
anova(fit)

# diagnostic plots
layout(matrix(c(1,2,3,4),2,2)) # optional layout
plot(fit) # diagnostic plots

# normality again
shapiro.test(residuals(fit))

# independence
durbinWatsonTest(fit)
layout(matrix(c(1),1,1))
acf(residuals(fit))

# nice wrapper function to generally test a lot of stuff
gvmodel <- gvlma(fit)
summary(gvmodel)


```

No significant differences.

### Post Weight

Test for significant differences after intervention

```{r}

# Post INtervention Weight - should beno difference
# not controls
fit <- lm(Weight_post ~ Intervention, mydata)
anova(fit)
summary(fit)

fit <- lm(Weight_post ~ Intervention + c.age + hispanic + c.stress,
          mydata)
anova(fit)
summary(fit)

# diagnostic plots
layout(matrix(c(1,2,3,4),2,2)) # optional layout
plot(fit) # diagnostic plots

# normality again
shapiro.test(residuals(fit))

# independence
durbinWatsonTest(fit)
layout(matrix(c(1),1,1))
acf(residuals(fit))

# nice wrapper function to generally test a lot of stuff
gvmodel <- gvlma(fit)
summary(gvmodel)


```

### Difference Scores of Weight

Test for significant differences after intervention

```{r}

# Post INtervention Weight - should beno difference
# not controls
fit <- lm(Weight_diff ~ Intervention, mydata)
anova(fit)
summary(fit)

fit <- lm(Weight_diff ~ Intervention + c.age + hispanic + c.stress,
          mydata)
anova(fit)
summary(fit)
results.diff[["Weight"]] <- summary(fit)[["coefficients"]]

# diagnostic plots
layout(matrix(c(1,2,3,4),2,2)) # optional layout
plot(fit) # diagnostic plots

# normality again
shapiro.test(residuals(fit))

# independence
durbinWatsonTest(fit)
layout(matrix(c(1),1,1))
acf(residuals(fit))

# nice wrapper function to generally test a lot of stuff
gvmodel <- gvlma(fit)
summary(gvmodel)


```

Therefore, we have evidence that weight (kg) is not effected by Prebiotic vs. placebo.


## Results Adjusted for Multiple Comparisons

```{r echo=F}

unlist.res <- function(dat, ...){
 dat<- results.diff
  nc <- ncol(dat[[1]])
  cnm <- colnames(dat[[1]])
  rnm <- names(dat)
  
  # unlist
  res <- data.frame(matrix(unlist(dat[[1]]), ncol=nc))
  colnames(res) <- colnames(dat[[1]])
  res$Outcome <- rep(rnm[1], nrow(res))
  res$Parameter <- rownames(dat[[1]])
  for(i in 2:length(rnm)){
    resi <- data.frame(matrix(unlist(dat[[i]]), ncol=nc))
    colnames(resi) <- colnames(dat[[i]])
    resi$Outcome <- rep(rnm[i], nrow(resi))
    resi$Parameter <- rownames(dat[[i]])
    res <- full_join(res, resi)
  }
  
  return(res)
}


```


```{r}


all_results <- unlist.res(results.diff)

all_results$p.adj <- p.adjust(all_results$`Pr(>|t|)`, "fdr") 
all_results$sigflag <- ifelse(all_results$p.adj < 0.05, "*", " ")
all_results$`Pr(>|t|)` <- round(all_results$`Pr(>|t|)`, 3)
all_results$p.adj.fdr <- round(all_results$p.adj, 3)
all_results <- all_results[,c(5:6, 1:4,7:8)]

kable(all_results, format="html", digits=3) %>%
  kable_styling(full_width = T)

```



