---
title: "Beta Diversity Analysis"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

# Beta-Diversity



```{r data, include=FALSE, echo=FALSE}
# load packages
source("code/load_packages.R")

# get data
source("code/get_cleaned_data.R")

theme_set(theme_bw())
pal = "Set1"
scale_colour_discrete <-  function(palname=pal, ...){
  scale_colour_brewer(palette=palname, ...)
}
scale_fill_discrete <-  function(palname=pal, ...){
  scale_fill_brewer(palette=palname, ...)
}

knitr::opts_chunk$set(out.width = "200%")


# recode
plot_data <- microbiome_data$meta.dat
plot_data$Week <- factor(plot_data$Week,
                         levels=c("1", "4", "8", "12"),
                         labels=c("Week 1", "Week 4",
                                  "Week 8", "Week 12"))

so <- distinct(microbiome_data$meta.dat, SubjectID, .keep_all = T)

subjectorder <- so$SubjectID[order(so$Intervention, decreasing = F)]

plot_data$SubjectID10 <- factor(plot_data$SubjectID,
                             levels = subjectorder,
                             labels=c(1:11))

ids <- 1:6
plot_data$Intervention2 <- ifelse(plot_data$SubjectID10 %in% ids, "Group A", "Group B")


# color pallet
cols <- c("#00a2f2",  "#c91acb", "#7f5940", "#cc5200", "#00d957", "#40202d", "#e60099", "#006fa6", "#f29d3d", "#300059", "#566573", "#336655", "#83008c", "#d9a3aa", "#400009", "#0020f2", "#a3d936", "#8091ff", "#fbffbf", "#00ffcc", "#8c4f46", "#354020", "#39c3e6", "#333a66", "#ff0000", "#6a8040", "#a6538a", "#402910", "#730f00", "#0a4d00", "#ffe1bf", "#a3d9b1", "#003033", "#f29979", "#00b3a7", "#cbace6", "#bfd9ff", "#bf0000", "#293aa6", "#594943", "#e5c339")




```


# General Ordination Plot

```{r}

# Ordinate
pcoa <- ordinate(
  physeq = phylo_data, 
  method = "PCoA", 
  distance = "bray"
)

# Plot 
p <- plot_ordination(
  physeq = phylo_data,
  ordination = pcoa,
  color = "Week",
  shape = "Intervention",
  title = "PCoA plot of community differences over time"
) 
p +
  geom_point(aes(color = Week), alpha = 0.7, size = 4) +
  scale_color_manual(values = c("#a65628", "red", "#ffae19","#4daf4a")) +
  geom_point(colour = "grey90", size = 1.5) 

```




# PERMANOVA

Below, we present the permutation based ANOVA results for community differences.
We conducted these analyses with three models

1. Using Week
2. Using Intervention
3. Using Week $\times$ Intervention interaction

Compute the distances matrices and get ordination objects.

```{r}

# Calculate distance matrices
dist_bray <- phyloseq::distance(phylo_data, method = "bray")
dist_unwt <- phyloseq::distance(phylo_data, method="unifrac", weighted=F)
dist_wt <- phyloseq::distance(phylo_data, method="unifrac", weighted=T)

# plot ordination
ord_bray = ordinate(phylo_data, method="PCoA", distance=dist_bray)
ord_unwt = ordinate(phylo_data, method="PCoA", distance=dist_unwt)
ord_wt = ordinate(phylo_data, method="PCoA", distance=dist_wt)

# sample data
df <- data.frame(sample_data(phylo_data))

# colors
cols <- c("#fe9700","#00a2f2", "#662a00", "#c91acb","grey60","#858c69", "#a8863a", "#737373", "#d43f1f", "#5dd047",  "#ffff59")

```

## Analysis by Week

```{r fig1, fig.height = 8, fig.width = 4, fig.align = "center"}

p1 <- plot_ordination(phylo_data, ord_bray, color="Week") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  labs(title="PCoA on Bray-Curtis distance, by Week")

p2 <- plot_ordination(phylo_data, ord_unwt, color="Week") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  labs(title="PCoA on Unweighted Unifrac distance, by Week")

p3 <- plot_ordination(phylo_data, ord_wt, color="Week") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  labs(title="PCoA on Weight Unifrac distance, by Week")
p1 + p2 + p3 + plot_layout(ncol=1)

# Bray-Curtis
adonis(dist_bray ~ Week, data = df)
beta <- betadisper(dist_bray, df$Week)
permutest(beta)

# Unweighted Unifrac
adonis(dist_unwt ~ Week, data = df)
beta <- betadisper(dist_unwt, df$Week)
permutest(beta)

# Weighted Unifrac
adonis(dist_wt ~ Week, data = df)
beta <- betadisper(dist_wt, df$Week)
permutest(beta)

```

## Analysis by Intervention

```{r fig2, fig.height = 8, fig.width = 4, fig.align = "center"}


p1 <- plot_ordination(phylo_data, ord_bray, color="Intervention") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  labs(title="PCoA on Bray-Curtis distance, by Intervention")

p2 <- plot_ordination(phylo_data, ord_unwt, color="Intervention") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  labs(title="PCoA on Unweighted Unifrac distance, by Intervention")

p3 <- plot_ordination(phylo_data, ord_wt, color="Intervention") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  labs(title="PCoA on Weight Unifrac distance, by Intervention")
p1 + p2 + p3 + plot_layout(ncol=1)

# Bray-Curtis
adonis(dist_bray ~ Intervention, data = df)
beta <- betadisper(dist_bray, df$Intervention)
permutest(beta)

# Unweighted Unifrac
adonis(dist_unwt ~ Intervention, data = df)
beta <- betadisper(dist_unwt, df$Intervention)
permutest(beta)

# Weighted Unifrac
adonis(dist_wt ~ Intervention, data = df)
beta <- betadisper(dist_wt, df$Intervention)
permutest(beta)

```


## Analysis by Intervention

```{r fig3, fig.height = 8, fig.width = 4, fig.align = "center"}

p1 <- plot_ordination(phylo_data, ord_bray,
                      color="Week", shape="Intervention") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  geom_point(colour = "grey90", size = 1.5) +
  labs(title="PCoA on Bray-Curtis distance, by Week & Intervention")+
  theme(legend.position = "none")

p2 <- plot_ordination(phylo_data, ord_unwt,
                      color="Week", shape="Intervention") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  geom_point(colour = "grey90", size = 1.5) +
  labs(title="PCoA on Unweighted Unifrac distance, by Week & Intervention")+
  theme(legend.position = "none")

p3 <- plot_ordination(phylo_data, ord_wt,
                      color="Week", shape="Intervention") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  geom_point(colour = "grey90", size = 1.5) +
  labs(title="PCoA on Weight Unifrac distance, by Week & Intervention")+
  theme(legend.position = "bottom")
p1 + p2 + p3 + plot_layout(ncol=1)

# Bray-Curtis
adonis(dist_bray ~ Intervention*Week, data = df)
beta <- betadisper(dist_bray, interaction(df$Intervention, df$Week))
permutest(beta)

# Unweighted Unifrac
adonis(dist_unwt ~ Intervention*Week, data = df)
beta <- betadisper(dist_unwt, interaction(df$Intervention, df$Week))
permutest(beta)

# Weighted Unifrac
adonis(dist_wt ~ Intervention*Week, data = df)
beta <- betadisper(dist_wt, interaction(df$Intervention, df$Week))
permutest(beta)

```



# Beta diversity plot (Figure 2)


```{r}

p1 <- plot_ordination(phylo_data, ord_bray, color="Week") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  labs(title="PCoA on Bray-Curtis distance, by Week")

p2 <- plot_ordination(phylo_data, ord_wt, color="Week") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  labs(title="PCoA on Weight Unifrac distance, by Week")


dat.bray <- p1$data
dat.wt <- p2$data

so <- distinct(microbiome_data$meta.dat, SubjectID, .keep_all = T)

subjectorder <- so$SubjectID[order(so$Intervention, decreasing = F)]

dat.bray$SubjectID10 <- factor(dat.bray$SubjectID,
                             levels = subjectorder,
                             labels=c(1:11))


dat.wt$SubjectID10 <- factor(dat.wt$SubjectID,
                             levels = subjectorder,
                             labels=c(1:11))

ids <- 1:6
dat.bray$Intervention2 <- ifelse(dat.bray$SubjectID10 %in% ids, "Group A", "Group B")
dat.wt$Intervention2 <- ifelse(dat.wt$SubjectID10 %in% ids, "Group A", "Group B")


p1 <- ggplot(dat.bray, aes(x=Week, y=Axis.1, color=Intervention2)) +
  geom_point(size=3)+
  facet_wrap(.~SubjectID10,ncol=6) +
  scale_color_manual(values=cols)+
  labs(y="PC1", 
       title="Individualized beta-diversity over study diration")+
  guides(colour= guide_legend(title = "Intervention"))+
  theme(panel.grid = element_blank(),
        strip.text.x = element_text(angle = 0, size = 11, face = "italic"),
        axis.text.y = element_text(size = 10),
        axis.title = element_text(size = 10),
        plot.title = element_text(hjust = 0.5),
        #axis.title.x = element_blank(),
        strip.background = element_blank(),
        legend.position = "right",
        #legend.text = element_text(size = 7),
        #legend.title = element_blank(),
        panel.spacing.x=unit(0.001, "lines"))

p1

  library(cowplot)
  p1.leg <- get_legend(p1)
  
  p1.nl <- p1 + theme(legend.position = "none")

  ggsave("fig/figure2.pdf", p1.nl, width=7.9,height=4.5, units="in")
  
  
  bigplotlegend <- plot_grid(p1.leg, nrow =1)
  
  save_plot("fig/figure2_legend.pdf", bigplotlegend,
            base_width = 7.9, base_height = 4.5)
  
```


# Beta diversity plot (Figure 3)


```{r}

p1 <- plot_ordination(phylo_data, ord_bray,
                      color="Week", shape="Intervention") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  labs(title="PCoA on Bray-Curtis distance",
       tag="A")+
  theme(panel.grid = element_blank(),
        axis.title = element_text(size = 10),
        plot.title = element_text(hjust = 0.5))

p2 <- plot_ordination(phylo_data, ord_unwt,
                      color="Week", shape="Intervention") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  labs(title="PCoA on Unweighted Unifrac distance",
       tag="B")+
  theme(panel.grid = element_blank(),
        axis.title = element_text(size = 10),
        plot.title = element_text(hjust = 0.5))

p3 <- plot_ordination(phylo_data, ord_wt,
                      color="Week", shape="Intervention") +
  geom_point(size=3) +
  scale_colour_manual(values=cols)+
  labs(title="PCoA on Weight Unifrac distance",
       tag="C")+
  theme(panel.grid = element_blank(),
        axis.title = element_text(size = 10),
        plot.title = element_text(hjust = 0.5))


  library(cowplot)
  p1.leg <- get_legend(p1)
  
  p1.nl <- p1 + theme(legend.position = "none")
  p2.nl <- p2 + theme(legend.position = "none")
  p3.nl <- p3 + theme(legend.position = "none")
  p <- p1.nl + p2.nl + p3.nl + plot_layout(ncol=1)
  ggsave("fig/figure3.pdf", p, width=4,height=6, units="in")
  
  
  bigplotlegend <- plot_grid(p1.leg, nrow =1)
  
  save_plot("fig/figure3_legend.pdf", bigplotlegend,
            base_width = 7.9, base_height = 4.5)
  
```
