---
title: "Alpha Diversity"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

This page contains the investigation of alpha diversity investigations.

```{r data, include=FALSE, echo=FALSE}
##Chunk iptions

# load packages
source("code/load_packages.R")

# get data
source("code/get_cleaned_data.R")

theme_set(theme_bw())
knitr::opts_chunk$set(out.width = "225%")

# recode
plot_data <- microbiome_data$meta.dat
plot_data$Week <- factor(plot_data$Week,
                         levels=c("1", "4", "8", "12"),
                         labels=c("Week 1", "Week 4",
                                  "Week 8", "Week 12"))

so <- distinct(microbiome_data$meta.dat, SubjectID, .keep_all = T)

subjectorder <- so$SubjectID[order(so$Intervention, decreasing = F)]

plot_data$SubjectID10 <- factor(plot_data$SubjectID,
                             levels = subjectorder,
                             labels=c(1:11))

ids <- 1:6
plot_data$Intervention2 <- ifelse(plot_data$SubjectID10 %in% ids, "Group A", "Group B")


# color pallet
cols <- c("#00a2f2",  "#c91acb", "#7f5940", "#cc5200", "#00d957", "#40202d", "#e60099", "#006fa6", "#f29d3d", "#300059", "#566573", "#336655", "#83008c", "#d9a3aa", "#400009", "#0020f2", "#a3d936", "#8091ff", "#fbffbf", "#00ffcc", "#8c4f46", "#354020", "#39c3e6", "#333a66", "#ff0000", "#6a8040", "#a6538a", "#402910", "#730f00", "#0a4d00", "#ffe1bf", "#a3d9b1", "#003033", "#f29979", "#00b3a7", "#cbace6", "#bfd9ff", "#bf0000", "#293aa6", "#594943", "#e5c339")


```


# Alpha diversity

## Boxplots by unique participants


### Observed OTUs

```{r}



p <- ggplot(plot_data, aes(x=SubjectID10, y=Observed,
                      colour=SubjectID10, shape=Week))+
  geom_point(size=3) +
  facet_grid(.~Intervention2, scales = "free")+
  labs(x="Subject ID", y= "Observed OTUs")+
  guides(colour= guide_legend(title = "Subject ID"))+
  theme(panel.grid = element_blank(),
        strip.text.x = element_text(angle = 0, size = 11, face = "italic"),
        axis.text.y = element_text(size = 10),
        axis.title = element_text(size = 10),
        plot.title = element_text(hjust = 0.5),
        axis.title.x = element_blank(),
        strip.background = element_blank(),
        legend.position = "right",
        #legend.text = element_text(size = 7),
        #legend.title = element_blank(),
        panel.spacing.x=unit(0.001, "lines"))
p
#ggsave("fig/figure1.pdf", p, width=5, height=3.5, units="in")



```

### Shannon Index

```{r}

ggplot(plot_data, aes(x=SubjectID3, y=Shannon,
                      colour=SubjectID3, shape=Week))+
  geom_point(size=2) +
  labs(x="Subject ID")+
  guides(colour= guide_legend(title = "Subject ID"))+
  theme(axis.text.x = element_text(angle=90, vjust=0.5))


```

### Simpson

```{r}

ggplot(plot_data, aes(x=SubjectID3, y=Simpson,
                      colour=SubjectID3, shape=Week))+
  geom_point(size=2) +
  labs(x="Subject ID")+
  guides(colour= guide_legend(title = "Subject ID"))+
  theme(axis.text.x = element_text(angle=90, vjust=0.5))

```



## Boxplots by week of study

### Observed OTUs

```{r}

ggplot(plot_data, aes(x=Week, y=Observed,
                      colour=SubjectID3))+
  geom_boxplot(outlier.alpha = 0,aes(group=Week))+
  geom_point(size=2, position="jitter")+
  labs(x="Week of Study")+
  guides(colour= guide_legend(title = "Subject ID"))+
  theme(axis.text.x = element_text(angle = 0, hjust=0.33))


```

### Shannon Index

```{r}

ggplot(plot_data, aes(x=Week, y=Shannon, colour=SubjectID3))+
  geom_boxplot(outlier.alpha = 0,aes(group=Week))+
  geom_point(size=2, position="jitter")+
  labs(x="Week of Study", y="Shannon Index")+
  guides(colour= guide_legend(title = "Subject ID"))+
  theme(axis.text.x = element_text(angle = 0, hjust=0.33))


```

### Simpson

```{r}

ggplot(plot_data, aes(x=Week, y=Simpson, colour=SubjectID3))+
  geom_boxplot(outlier.alpha = 0,
               aes(group=Week))+
  geom_point(size=2, position="jitter")+
  labs(x="Week of Study", y="Simpson")+
  guides(colour= guide_legend(title = "Subject ID"))+
  theme(axis.text.x = element_text(angle = 0, hjust=0.33))


```


## Manuscript Figure

```{r}

p1 <- ggplot(plot_data, aes(x=SubjectID10, y=Observed,
                      colour=SubjectID10, shape=Week))+
  geom_point(size=3) +
  facet_grid(.~Intervention2, scales = "free")+
  labs(x="Subject ID", y= "Observed OTUs", tag="A",
       title="Observed OTUs by Participants")+
  guides(colour= guide_legend(title = "Subject ID"))+
  scale_colour_manual(values = cols)+
  theme(panel.grid = element_blank(),
        strip.text.x = element_text(angle = 0, size = 11, face = "italic"),
        axis.text.y = element_text(size = 10),
        axis.title = element_text(size = 10),
        plot.title = element_text(hjust = 0.5),
        axis.title.x = element_blank(),
        strip.background = element_blank(),
        legend.position = "right",
        #legend.text = element_text(size = 7),
        #legend.title = element_blank(),
        panel.spacing.x=unit(0.001, "lines"))
p1

p2 <- ggplot(plot_data, aes(x=Intervention, y=Observed,
                            colour=Intervention,
                            group=Intervention))+
  geom_boxplot(outlier.shape=NA)+
  geom_point(size=3, position="jitter")+
  facet_wrap(.~Week, nrow=1)+
  labs(x=NULL, y="Observed OTUs",
       tag="B", title="Observed OTUs by Study Week")+
  guides(colour= guide_legend(title = "Intervention"))+
  scale_colour_manual(values = cols)+
  theme(panel.grid = element_blank(),
        strip.text.x = element_text(angle = 0, size = 11, face = "italic"),
        axis.text.y = element_text(size = 10),
        axis.title = element_text(size = 10),
        plot.title = element_text(hjust = 0.5),
        axis.title.x = element_blank(),
        strip.background = element_blank(),
        legend.position = "right",
        #legend.text = element_text(size = 7),
        #legend.title = element_blank(),
        panel.spacing.x=unit(0.001, "lines"))
p2 

  library(cowplot)
  p1.leg <- get_legend(p1)
  p2.leg <- get_legend(p2)
  
  p1.nl <- p1 + theme(legend.position = "none")
  p2.nl <- p2 + theme(legend.position = "none")
  
  p <- p1.nl + p2.nl + plot_layout(nrow=1) 
  p
  ggsave("fig/figure1.pdf", p, width=7.9,height=5, units="in")
  
  
  bigplotlegend <- plot_grid(p1.leg, p2.leg, nrow =1, align = "h")
  
  save_plot("fig/figure1_legend.pdf", bigplotlegend, base_width = 13.25, base_height = 5)

```



# Intervention Effect

## Overall Effect

### Observed OTUs

```{r}

ggplot(plot_data, aes(x=Intervention, y=Observed))+
  geom_boxplot(outlier.alpha = 0,
               aes(group=Intervention, color=Intervention))+
  geom_point(position="jitter",aes(color=Intervention))+
  guides(colour= guide_legend(title = "Intervention"))+
  labs(x="Intervention Group")+
  theme(axis.text.x = element_text(angle = 0, hjust=0.33))


```

### Shannon Index

```{r}

ggplot(plot_data, aes(x=Intervention, y=Shannon))+
  geom_boxplot(outlier.alpha = 0,
               aes(group=Intervention, color=Intervention))+
  geom_point(position="jitter",aes(color=Intervention))+
  guides(colour= guide_legend(title = "Intervention"))+
  labs(x="Intervention Group")+
  theme(axis.text.x = element_text(angle = 0, hjust=0.33))


```

### Simpson

```{r}

ggplot(plot_data, aes(x=Intervention, y=Simpson))+
  geom_boxplot(outlier.alpha = 0,aes(group=Intervention, color=Intervention))+
  geom_point(position="jitter",aes(color=Intervention))+
  guides(colour= guide_legend(title = "Intervention"))+
  labs(x="Intervention Group")+
  theme(axis.text.x = element_text(angle = 0, hjust=0.33))


```

## Changes Over Time

### Observed OTUs

```{r}

ggplot(plot_data, aes(x=Intervention, y=Observed,
                      group=Intervention,
                      color=Intervention, shape=Intervention))+
  geom_boxplot(outlier.alpha = 0)+
  geom_point(position="jitter")+
  facet_grid(.~Week)+
  labs(x="Intervention Group by Week")+
  guides(colour= guide_legend(title = "Intervention"))+
  theme(axis.text.x = element_text(angle = 0, hjust=0.33))

ggplot(plot_data, aes(x=Week, y=Observed,
                      group=interaction(Intervention, SubjectID3),
                      color=SubjectID3, shape=Intervention))+
  #geom_boxplot(outlier.alpha = 0)+
  geom_point()+
  geom_line()+
  #facet_grid(.~Week)+
  labs(x="Intervention Group by Week")+
  guides(colour= guide_legend(title = "Intervention"))+
  theme(axis.text.x = element_text(angle = 0, hjust=0.33))

```

### Shannon Index

```{r}

ggplot(plot_data, aes(x=Intervention, y=Shannon,
                      group=Intervention,
                      color=Intervention, shape=Intervention))+
  geom_boxplot(outlier.alpha = 0)+
  geom_point(position="jitter")+
  facet_grid(.~Week)+
  labs(x="Intervention Group by Week")+
  guides(colour= guide_legend(title = "Intervention"))+
  theme(axis.text.x = element_text(angle = 0, hjust=0.33))


```

### Simpson

```{r}

ggplot(plot_data, aes(x=Intervention, y=Simpson,
                      group=Intervention,
                      color=Intervention, shape=Intervention))+
  geom_boxplot(outlier.alpha = 0)+
  geom_point(position="jitter")+
  facet_grid(.~Week)+
  labs(x="Intervention Group by Week")+
  guides(colour= guide_legend(title = "Intervention"))+
  theme(axis.text.x = element_text(angle = 0, hjust=0.33))


```

