---
title: "Summary Statistics"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

This page contains the summary statistics of participants' data across the study.

```{r}
# load packages
source("code/load_packages.R")

# get data
source("code/get_cleaned_data.R")

```


# Participant Demographics

## Summarize by Week

```{r}

mydata <- microbiome_data$meta.dat %>%
  mutate(female = ifelse(Gender == "F", 1, 0),
         c.age = Age - mean(Age),
         IntB = ifelse(Intervention == "B", 1, 0),
         Stress = Stress.Scale,
         hispanic = ifelse(Ethnicity %in% c("White", "Asian", "Native America"), 1, 0),
         BMI = Weight_pre/((Height_cm/100)**2))

sumTab <- mydata %>%
  dplyr::group_by(Week) %>%
  dplyr::summarise(N = n(),
            Age_Mean = mean(Age),
            Age_SD = sd(Age),
            Weight_kg_M = mean(Weight_kg),
            Weight_kg_SD = sd(Weight_kg),
            Height_cm_M = mean(Height_cm),
            Height_cm_SD = sd(Height_cm))

kable(sumTab, format="html", digits=1) %>%
  kable_styling(full_width = T)

```


## Baseline Only for Significance Test


```{r}

sumTab <- mydata %>%
  dplyr::filter(Week==1) %>%
  dplyr::group_by(Intervention) %>%
  dplyr::summarise(N = n(),
            PercentMale = (1 - mean(female))*100,
            Age_Mean = mean(Age),
            Age_SD = sd(Age),
            Weight_kg_M = mean(Weight_kg),
            Weight_kg_SD = sd(Weight_kg),
            Height_cm_M = mean(Height_cm),
            Height_cm_SD = sd(Height_cm),
            BMI_M = mean(BMI),
            BMI_SD = sd(BMI),
            VFL_M = mean(Visceral_Fat_Level_pre),
            VFL_SD = sd(Visceral_Fat_Level_pre),
            PercBF_M = mean(Perc_Body_Fat_pre),
            PercBF_SD = sd(Perc_Body_Fat_pre),
            PercLF_M = mean(LBM_pre),
            PercLF_SD = sd(LBM_pre),
            HEI_Total_M = mean(HEI_Total, na.rm=T),
            HEI_Total_SD = sd(HEI_Total, na.rm=T))

sumTab <- t(sumTab)
M <- sumTab[rownames(sumTab) %like% "_M",]
SD <- sumTab[rownames(sumTab) %like% "_SD",]
tab <- cbind(M, SD)
tab <- tab[, c(1,3,2,4)]
Ng <- c(sumTab[2,1], NA, sumTab[2,2], NA)
PercentMale <- c(sumTab[3,1], NA, sumTab[3,2], NA)

tab <- rbind(Ng, PercentMale, tab)
colnames(tab) <- c("GroupA_Mean","GroupA_SD", "GroupB_Mean", "GroupB_SD")

tab <- apply(tab, 1:2, as.numeric)
#tab
```

Next, we conducted Mann-Whitney-U test to compare distributions across intervention vs. placebo groups.


```{r}

dat <- filter(mydata, Week == 1)

VAR <- c("Age", "Weight_kg", "Height_cm", "BMI", "Visceral_Fat_Level_pre", "Perc_Body_Fat_pre", "LBM_pre", "HEI_Total")
out <- numeric(length(VAR))

i <- 1
for(i in 1:length(VAR)){
  fit <- wilcox.test(dat[,VAR[i]] ~ dat$IntB)
  out[i] <- fit$p.value
}

out <- c(NA, NA, out)
out <- matrix(out, ncol=1)
colnames(out) <- "P_Value"
tab <- cbind(tab, out)

kable(tab, format="html", digits=3) %>%
  kable_styling(full_width = T)
tab <- as.data.frame(tab)
#write.csv(tab,paste0(w.d, "/tab/table_1_results.csv"))

```

## numbers by ASA24, FFQ, Stool, and Blood Samples


### ASA24 

```{r}


asa24 <- microbiome_data$meta.dat

# If the recall number is missing, then they didn't response so we need to exclude them in the count.

asa24 %>%
  dplyr::group_by(Week)%>%
  filter(is.na(RecallNo)==F)%>%
  dplyr::summarise(N=n())

```

### Food Frequency Questionnaire (FFQ)

HEI scores were only gathered at onetime point.
So, we need to subset to only 1 week then exclude the missing observations.


```{r}

ffq <- microbiome_data$meta.dat

ffq %>%
  dplyr::filter(Week == 1, is.na(HEI_Total)==F)%>%
  dplyr::summarise(N = n())

```

### Stool Samples

```{r}

Nvec <- matrix(ncol=1, nrow=5, dimnames = list(c("Total", "Week1", "Week4", "Week8", "Week12"), c("N")))

# total number of samples
Nvec[1,1] <- nsamples(phylo_data)

# week 1
subph <- subset_samples(phylo_data, Week == 1)
Nvec[2,1] <- nsamples(subph)

# week 4
subph <- subset_samples(phylo_data, Week == 4)
Nvec[3,1] <- nsamples(subph)

# week 8
subph <- subset_samples(phylo_data, Week == 8)
Nvec[4,1] <- nsamples(subph)

# week 12
subph <- subset_samples(phylo_data, Week == 12)
Nvec[5,1] <- nsamples(subph)

Nvec

```


### Blood Samples

```{r}
# Blood Samples
blood_data <- read_excel("data/Blood measures data/Copy of Fiber Study Blood Results.xlsx")

meta_data <- microbiome_data$meta.dat

keepVar <- c("SubjectID","Week", "Intervention", "Stress.Scale", "Ethnicity", "Gender", "Age")
meta_data <- meta_data [, keepVar] %>%
  filter(Week == 1)

blood_data <- full_join(blood_data, meta_data)

blood_data <- distinct(blood_data, SubjectID, time,.keep_all = T)

blood_data %>%
  dplyr::group_by(time) %>%
  dplyr::summarise(N=n())

```


